Clarifying Correlation and Causation: A Guide to Modern Quantitative Causal Inference in Nonprofit Studies

### 2024-11-08 manuscript/manuscript.qmd
Add outline for ARNOVA paper

---
title: "Clarifying Correlation and Causation: A Guide to Modern Quantitative Causal Inference in Nonprofit Studies"
short-title: "Clarifying Correlation and Causation"
date: "December 14, 2023"
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
- name: Meng Ye
  affiliation: Georgia State University
  email: mye2@gsu.edu
  url: https://aysps.gsu.edu/phd-student/ye-meng/
  orcid: 0000-0002-6552-8338
correspondence-prefix: "Correspondence concerning this article should be addressed to"
published: Drafting
abstract: |
  Discovering causal relationships and testing theoretical mechanisms is a core endeavor of social science. Randomized experiments have long served as a gold standard for making valid causal inferences, but most of the data social scientists work with is observational and non-experimental. However, with newer methodological developments in economics, political science, epidemiology, and other disciplines, an increasing number of studies in social science make causal claims with observational data. As a newer interdisciplinary field, however, nonprofit studies has lagged behind other disciplines in its use of observational causal inference. In this paper, we present a hands-on introduction and guide to design-based observational causal inference methods. We first review and categorize all studies making causal claims in top nonprofit studies journals over the past decade to illustrate the field’s current of experimental and observational approaches to causal inference. We then introduce a framework for modeling and identifying causal processes using directed acyclic graphs (DAGs) and provide a walk-through of the assumptions and procedures for making inferences with a range of different methods, including matching, inverse probability weighting, difference-in-differences, regression discontinuity designs, and instrumental variables. We illustrate each approach with synthetic and empirical examples and provide sample R and Stata code for implementing these methods. We conclude by encouraging scholars and practitioners to make more careful and explicit causal claims in their observational empirical research, collectively developing and improving quantitative work in the broader field of nonprofit studies.
thanks: |
  Acknowledgments here if any
epigraph:
  - text: "Don't let us forget that the causes of human actions are usually immeasurably more complex and varied than our subsequent explanations of them."
    source: "Fyodor Dostoevsky, *The Idiot*, part IV, chapter 2"
keywords:
  - those
  - go
  - here
word-count: ""
bibliography: references.json
link-citations: true
---

@@

options(
  digits = 3, width = 120,
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)
```

@@

>>>>

---
title: "Clarifying Correlation and Causation: A Guide to Modern Quantitative Causal Inference in Nonprofit Studies"
short-title: "Clarifying Correlation and Causation"
date: 2024-11-08
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
- name: Meng Ye
  affiliation: Georgia State University
  email: mye2@gsu.edu
  url: https://aysps.gsu.edu/phd-student/ye-meng/
  orcid: 0000-0002-6552-8338
correspondence-prefix: "Correspondence concerning this article should be addressed to"
published: Drafting
abstract: |
  Discovering causal relationships and testing theoretical mechanisms is a core endeavor of social science. Randomized experiments have long served as a gold standard for making valid causal inferences, but most of the data social scientists work with is observational and non-experimental. However, with newer methodological developments in economics, political science, epidemiology, and other disciplines, an increasing number of studies in social science make causal claims with observational data. As a newer interdisciplinary field, however, nonprofit studies has lagged behind other disciplines in its use of observational causal inference. In this paper, we present a hands-on introduction and guide to design-based observational causal inference methods. We first review and categorize all studies making causal claims in top nonprofit studies journals over the past decade to illustrate the field’s current of experimental and observational approaches to causal inference. We then introduce a framework for modeling and identifying causal processes using directed acyclic graphs (DAGs) and provide a walk-through of the assumptions and procedures for making inferences with a range of different methods, including matching, inverse probability weighting, difference-in-differences, regression discontinuity designs, and instrumental variables. We illustrate each approach with synthetic and empirical examples and provide sample R and Stata code for implementing these methods. We conclude by encouraging scholars and practitioners to make more careful and explicit causal claims in their observational empirical research, collectively developing and improving quantitative work in the broader field of nonprofit studies.
thanks: |
  Acknowledgments here if any
epigraph:
  - text: "Don't let us forget that the causes of human actions are usually immeasurably more complex and varied than our subsequent explanations of them."
    source: "Fyodor Dostoevsky, *The Idiot*, part IV, chapter 2"
keywords:
  - those
  - go
  - here
word-count: ""
bibliography: references.json
link-citations: true
---

@@

options(
  digits = 3, width = 120
)
```

@@

::: {.content-visible when-format="html"}
# Introduction {.hide}
:::

::: {.callout-note}
# ARNOVA outline

Cleaned up, simplified outline that I'll expand and integrate with the rest of the outline/draft below
:::

Causal inference important in social sciences, but hard. Nobel prize for Angrist and Imbens and gang...

Answering causal questions is an important endeavor for nonprofit researchers and practitioners. Nonprofits serve social goals and create and implement a range of programs to accomplish those goals. Measuring the impact of nonprofit programming is essential both for grantors that require evidence of program impact and for nonprofit program managers who are interested in whether their programs lead to the desired outcomes.

Causal work in the nonprofit policy and management literature has increased in the past decade (see fig1), but it remains a less explored topic. The nonprofit literature is full of careful quantitative and statistical work, explicitly descriptive, implicitly causal (with weasel words), and explicitly causal.

In this paper, we provide a practical reference, guide, and review of modern causal inference methods. For the sake of accessibility, we provide sample data, R and Stata code, minimal—and extensively annotated—mathematical notation. We offer a unified, cross-disciplinary framework for thinking about causal identification, spanning econometrics and epidemiology, providing a toolbox for nonprofit researchers to think more carefully about causally-oriented research design and analysis, leading to more effective evaluations of programs, policies, and interventions.

Roadmap—in this paper we blah blah blah

# The fundamental problem of casual inference

Explain that here…

# Visualizing, identifying, and isolating causal relationships with DAGs

General explanation of DAGs and confounding/colliding/mediating and d-separation and do-calculus, etc.

# Isolating causal relationships

To identify a causal relationship in a DAG, we have to ensure that the arrow between the treatment and outcome nodes is isolated and that the treatment and outcome nodes are not linked through any other pathways

There are two general approaches for doing this:

- Circumstantial identification
- Adjustment-based identification

# Situation-based approaches

## RCTs

- Explanation + DAG
- Illustration of how to use it
- Review of existing nonprofit studies that use it
- Possible nonprofit research questions that could use it

## Diff-in-diff + TWFE

- Explanation + DAG
- Illustration of how to use it
- Review of existing nonprofit studies that use it
- Possible nonprofit research questions that could use it

## RDD

- Explanation + DAG
- Illustration of how to use it
- Review of existing nonprofit studies that use it
- Possible nonprofit research questions that could use it

## IV

- Explanation + DAG
- Illustration of how to use it
- Review of existing nonprofit studies that use it
- Possible nonprofit research questions that could use it

# Adjustment-based approaches

## Regression adjustment

- Explanation + DAG
- Illustration of how to use it
- Review of existing nonprofit studies that use it
- Possible nonprofit research questions that could use it

## IPW and g-computation

- Explanation + DAG
- Illustration of how to use it
- Review of existing nonprofit studies that use it
- Possible nonprofit research questions that could use it

# Hybrid approaches

Briefly describe combinations of the two approaches, like synthetic controls, DiD with matching, RDD with covariate adjustment, etc.

# Tools for researchers and practitioners

# Conclusion

---
